---
title: "Visualising COVID data using R and leaflet"
subtitle: "Data visualisation series (Post #2)"
summary: "Data visualisation using R and leaflet"
author: "royr2"
date: 2021-09-12
categories: ["R", "Data Visualisation", "leaflet"]
tags: ["R", "leaflet", "data visualisation", "covid19"]  
comments: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

`Leaflet` is a JavaScript library for interactive maps. Leaflet for R is super easy to use and a great way to visualise data that has a spatial dimension. Below are some examples on how to use the `leaflet` package in R to visualise some COVID-19 data.

## Packages
```{r eval = F}
# Pacman is a package management tool 
install.packages("pacman")
```

```{r message = F, warning = F}
library(pacman)

# p_load automatically installs packages if needed
p_load(dplyr, leaflet)
```

## Sample dataset
A summarised version of the COVID-19 Data Repository hosted by [JHU](https://github.com/CSSEGISandData/COVID-19) is available for download [here](https://github.com/royr2/blog/blob/main/download/covid_data.csv)

```{r echo = F}
df <- read.csv("/Github/yetanothercovidapp/plot_data.csv")
```

```{r eval = F}
df <- read.csv("covid_data.csv")
```

## Confirmed cases across countries
Let's say we wanted to visualise how the cumulative case load is distributed across different countries on a map (as of a certain date).

```{r message = FALSE, warning = FALSE}
plt <- df %>% 
    filter(date == "2021-09-01") %>% 
    
    # Circle radius 
    # Arbitrary scaling function for dramatic effect
    mutate(rad = sqrt(confirmed/max(confirmed)) * 100) %>% 
    
    # Leaflet
    leaflet(options = leafletOptions(minZoom = 2, maxZoom = 6)) %>% 
    
    # Base map layer
    # Lots of other options see https://rstudio.github.io/leaflet/basemaps.html
    addProviderTiles(providers$Stamen.Toner,
                     options = providerTileOptions(opacity = 0.6)) %>%
    
    addCircleMarkers(lng = ~Longitude, 
                     lat = ~Latitude, 
                     radius = ~rad, 
                     popup = ~text,  ## Pop-ups understand html (see df$text)
                     weight = 0.7,
                     stroke =T, 
                     color = "#000000",
                     fillColor = "#79B4B7", 
                     fillOpacity = 0.3)
```

```{r echo = F}
htmlwidgets::saveWidget(plt, "map1.html")
```

```{r eval = F}
plt
```

<iframe src = "map1.html" height = "600" width = "900"></iframe>

The nice thing about leaflet is that visualisations are interactive (try clicking one of the bubbles). Note that the labels have already been generated. 

```{r}
head(df$text)
```

## Tracking confirmed cases over time
Now say we wanted to understand how the number of confirmed cases were distributed across countries at different points in time. Leaflet has a nice feature which allows users to add layers to the same graphic and UI elements (like radio buttons) to help switch between layers using the `group` argument and the `addLayersControl()` function. 

```{r message = FALSE, warning = FALSE}

# Date series (every 21 days)
vec <- seq.Date(as.Date("2020-03-01"), to = as.Date("2021-09-01"), by = "21 days")
vec <- as.character(vec)

plt <- df %>% 
    
    # Track the top 20 countries for each date
    group_by(date) %>% 
    
    arrange(desc(confirmed)) %>% 
    
    filter(row_number() <= 50, 
           date %in% vec) %>% 
    
    # Circle radius 
    # Arbitrary scaling function for dramatic effect
    mutate(rad = sqrt(confirmed/max(confirmed)) * 80) %>% 
    
    # Leaflet
    leaflet(options = leafletOptions(minZoom = 2, maxZoom = 6, )) %>% 
    
    # Base map layer
    # Lots of other options see https://rstudio.github.io/leaflet/basemaps.html
    addProviderTiles(providers$Stamen.TonerLite,
                     options = providerTileOptions(opacity = 0.8)) %>%
    
    addCircleMarkers(lng = ~Longitude, 
                     lat = ~Latitude, 
                     radius = ~rad, 
                     popup = ~text,
                     weight = 0.7,
                     stroke = T, 
                     color = "#000000",
                     fillColor = "#525c63", 
                     fillOpacity = 0.5, 
                     group = ~date, 
                     labelOptions = labelOptions(noHide = F)) %>% 
    # Layer control
    addLayersControl(
        
        # Using baseGroups adds radio buttons which makes it easier to switch
        baseGroups = vec,
        
        # Using overlayGroups adds checkboxes        
        # overlayGroups = vec
        
        options = layersControlOptions(collapsed = FALSE))
```

```{r echo = F}
htmlwidgets::saveWidget(plt, "map2.html")
```

```{r eval = F}
plt
```

<iframe src = "map2.html" height = "600" width = "900"></iframe>

It is interesting to note which countries were affected by COVID-19 in early 2020, how it then rapidly spread across Europe, the impact of lock-downs onthe number of confirmed cases, India's terrible second wave in mid 2021 and now the increasing number of cases in South East Asia. 

*Thoughts? Comments? Helpful? Not helpful? Like to see anything else added in here? Let me know!*

